Verification of Stencil Computations with Spatial Specifications

### 2016-10-15 new-intro.tex
Replace occurrences of (Sub)section with \S

In this paper, we make the following contributions:
%
\begin{itemize}
\item we introduce our specification language for stencils
  that captures many common forms of data access pattern
  (Section~\ref{sec:lang});

\item we detail checking and inference algorithms for our
  specifications (Section~\ref{sec:analysis}), derived from a
  denotational model for the specificaiton language
  (Section~\ref{sec:semantics});

@@

\item we report on a quantitative study of stencil computations on a
  corpus of numerical Fortran programs (ranging from small to large)
  (Section~\ref{sec:evaluation}), totalling one million lines of code. Our tool
  identifies and infers specifications for 35,000 stencil computations in the
  corpus. Approximately 10\% of the stencils we found are non-trivial,
  corresponding to code which is a possible source of errors. This validates
  our hypothesis that the majority of stencil computations have a regular
  shape, and validates the design of our language in its capability to capture
  many core patterns.

>>>>

In this paper, we make the following contributions:
%
\begin{itemize}
\item we introduce our specification language for stencils
  that captures many common forms of data access pattern
  (\S~\ref{sec:lang});

\item we detail checking and inference algorithms for our
  specifications (\S~\ref{sec:analysis}), derived from a
  denotational model for the specificaiton language
  (\S~\ref{sec:semantics});

@@

\item we report on a quantitative study of stencil computations on a
  corpus of numerical Fortran programs (ranging from small to large)
  (\S~\ref{sec:evaluation}), totalling one million lines of code. Our tool
  identifies and infers specifications for 35,000 stencil computations in the
  corpus. Approximately 10\% of the stencils we found are non-trivial,
  corresponding to code which is a possible source of errors. This validates
  our hypothesis that the majority of stencil computations have a regular
  shape, and validates the design of our language in its capability to capture
  many core patterns.

### 2016-10-15 paper.tex
revert to sigplan format
Replace occurrences of (Sub)section with \S

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{color}
\usepackage{graphics}
\usepackage{fancyvrb}
\usepackage{subfigure}
\usepackage{amsthm}
\usepackage{multirow}
\usepackage{minted}
\usepackage{breqn}
\usepackage{enumitem}
%\usepackage{subcaption}
\usepackage{siunitx} % For pretty-printing numeric values and SI units
                     % of measure. e.g., the tabular column type S is
                     % used to print nice-looking tables of numbers.
\sisetup{ % defaults
  group-separator={,},
  group-minimum-digits={3},
  output-decimal-marker={.},
  table-format = 6
}

\usepackage{tikz}
\usepackage{xypic}

@@

\newtheorem{lemma}[block]{Lemma}
\newtheorem{proposition}[block]{Proposition}
%\newtheorem{definition}[block]{Definisssstion}

@@

\title{Verification of Stencil Computations with Spatial
  Specifications}
\titlebanner{DRAFT -- Do not distribute}
%\authorinfo{}{}{}
\authorinfo{Dominic Orchard \and Mistral Contrastin
\and Matthew Danish \and Andrew Rice}{University of Cambridge}{}

@@

In this paper, we make the following contributions:
%
\begin{itemize}[leftmargin=0.8em]
\item we introduce a specification language for stencils 
  that captures many common forms of data access pattern
  (Section~\ref{sec:lang});

\item we detail inference, checking, and specification synthesis 
algorithms for our language (Section~\ref{sec:analysis}), derived from
a denotational model for the language (Section~\ref{sec:semantics});

@@

Our specification language is based on the hypothesis
that most forms of array access in numerical code have
a fixed, statically-determined access pattern. For example, the
\emph{five-point stencil} on a two-dimensional array reads from array
indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
for all $i, j$ within the inner boundary of the array (to avoid
out-of-bounds access at the edges). We revisit this hypothesis
in Section~\ref{sec:evaluation} with the inference of
such regular stencil patterns on our corpus of numerical programs.

Section~\ref{sec:syntax} outlines the syntax.
 Section~\ref{sec:eqs} provides an equational
theory for specifications and a notion of
sub-specification.  Section~\ref{sec:semantics}
gives a semantic model via a set-based interpretation, for which the
equality and sub-specification relation are proven sound.

@@

The last six equations (from (\textsc{\term{+}idem}) inclusive), explain
the algebraic behaviour of \term{+} and \term{*}.  Together, we see
that \term{+} is an associative, commutative, idempotent binary
operator that distributes with \term{*}, which is associative and
commutative. This distribution of \term{*} over \term{+} is key as it
is used internally to give a normal form for stencil specifications
akin to Disjunctive Normal Form (DNF) (where \term{+} is taken as
``disjunction'' and \term{*} as the ``conjunction''). We revisit this
normalisation in Section~\ref{sec:inf-step4}.

The equational theory provides not only an axiomatic semantics
for the language, which the user can use to rewrite their
specification, but it is also used extensively in the inference
procedure (Section~\ref{subsec:inference}) to simplify specifications.

@@

We define a denotational model of the semantics of our 
stencil specification language. This model has a number of purposes: (1) it is used in the inference,
checking, and synthesis algorithms (Section~\ref{sec:analysis}); (2)
justifies the equational
theory and approximations of the previous section; and (3) can be used
to guide correct implementations. The model domain is over sets of vectors which we call \emph{index
  schemes}, which we introduce first before the model in Section~\ref{subsec:model}.

@@

\begin{definition} \textsf{Mult}
and \textsf{Approx} are parametric labelled variant types
with injection function given by their definition:
%
\begin{align*}
\textsf{Mult} \, a & = \textsf{mult} \, a \,\mid\, \textsf{only} \, a \\
\textsf{Approx} \, a & = \textsf{exact} \, a \,\mid\, \textsf{lower} \, a \,\mid\,
\textsf{upper} \, a \,\mid\, \textsf{both} \, a \, a
\end{align*}
\ie{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$, and so
on. The $\textsf{Mult}$ type corresponds to the presence or absence of the
\term{readOnce} modifier as shown in Figure~\ref{fig:spatial-model}. 
The $\textsf{Approx}$ type corresponds to the presence or absence of
the spatial approximation modifier, with \textsf{exact} when there is
no such modifier and \textsf{lower} and \textsf{upper} for
\term{atLeast} and \term{atMost}. Later, it will be useful to model
pairs of lower and upper bounds which is provided by \textsf{both} 
(but which isn't used here). These are used for labelled models, and other components later in
Section~\ref{sec:analysis}.% with additional information.
\label{def:mult-and-approx}
\end{definition}
\vspace{-0.5em}

@@

\noindent
We briefly sketch the algorithms for inferring and synthesising
specifications (Section~\ref{subsec:inference}), for checking code
against specifications (Section~\ref{subsec:checking}). These utilise the semantic model of
Section~\ref{sec:semantics}.

@@

\begin{definition}\label{def:span-coalesc} Span
  coalescing $\bullet$ is defined for spans contiguous in the first dimension:
%%
\begin{align*}
\vect{x} \bullet \vect{y}
= &
\begin{cases}
(\vect{x}^L, \vect{y}^U) & \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
\vect{x}_{2:n} = \vect{y}_{2:n}) \\
\bot  & \textit{otherwise}
\end{cases}
% $m$-dimensions into coalesced spans which vary in
% $m+1$-dimensions.
\end{align*}
%%
\end{definition}
\noindent
To coalesce regions which are contiguous in other dimensions
other than the first, we generate all rotations of the spans in
$n$-dimensional space by permutation on spans, and fold together contiguous
regions using $\bullet$ (above). This 
generates all possible ways of coalescing regions. For our example,
the unit spans in Figure~\ref{fig:inference-steps-informal}(a)
are coalesced into the contiguous spans in
Figure~\ref{fig:inference-steps-informal}(b).
This procedure is iterated until a fixed point is reached. In this
example this is reached in the first step. 
 Then any spans that are completely contained by any other span are deleted,
leaving a minimal set of spans (which may overlap, but none of which
fully contains another), Figure~\ref{fig:inference-steps-informal}(c).
The final spans are used to synthesise concrete stencil specification
syntax (Section~\ref{sec:inf-step4}), where
each span becomes a product \texttt{*} of regions which are 
combined by the sum \texttt{+}.

@@

In checking mode, the access pattern of a stencil computation in the
source language is verified against a specification. The model of
Section~\ref{sec:semantics} again aids this process.  Checking
proceeds by generating a model $M$ from a specification and generating a set $N$ of indexing
schemes from the source code. The source code is then valid with
respect to the specification if $N$ is \emph{consistent} with $M$. The
notion of consistency is built out of index scheme 
\emph{generalisation} (Sect.~\ref{subsec:schemes-and-gen}). 
%Consistency is defined in Figure~\ref{fig:consistency},
We first set up some intermediate definitions.% before unpacking its definition.

@@

\begin{example}
We give an extended example of verifying a stencil computation against
its specification. We consider a more general, but still consistent, version of the \textit{five-point}
stencil specification inferred previously in
Section~\ref{subsec:inference}:
%
\begin{minted}[breakindent=8.1em]{fortran}
!= region :: fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
!= stencil readOnce, fivepoint :: b
a(i,j) = b(i,j-1)+b(i,j)+b(i,j+1)+b(i-1,j)+b(i+1,j)
\end{minted}
%
We refer to the assignment on line three as $l3$ for brevity.

Verification of the code against the spec is then given by 
$check(({\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}),l3,\texttt{b})$. 
We explain the steps.
\begin{enumerate}  
%
\item Generate the model from the specification as in
  Section~\ref{sec:semantics}. 
%
\begin{align*}
 M_s= \{  \vtwo{-1}{\infty},
          \vtwo{0}{\infty},
          \vtwo{1}{\infty},
          \vtwo{\infty}{-1},
          \vtwo{\infty}{0},
        \vtwo{\infty}{1} \}  \\[0.4em]
\interp{{\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}}_2 =
  \textsf{once}(\textsf{exact} (M_s))
\end{align*}
%
\item Obtained the code model via $\textit{modelise}$, giving:
%
\begin{align*}
M_c = \{
    \vtwo{0}{-1},
    \vtwo{0}{0},
    \vtwo{0}{1},
    \vtwo{-1}{0},
    \vtwo{1}{0}
  \}
\\[0.4em]
\textit{modelise}(l3, \texttt{b}) = \textsf{once} (M_c)
\end{align*}
%
\item Consistency $cons_2(\textsf{once}(\textsf{exact}(M_s)),
  \textsf{once}(M_c))$ holds when 
  $\textit{cons}_2(\textsf{mult}(\textsf{exact}(M_s)),\textsf{mult}(M_c))$ (Fig.~\ref{fig:consistency}) 
  which holds when: 
  \begin{align*}
     & \textit{cons}_2(\textsf{mult}(\textsf{up}(M_s)),\textsf{mult}(M_c)) \\
   \wedge \; & \textit{cons}_2(\textsf{mult}(\textsf{low}(M_s),\textsf{mult}(M_c))
  \end{align*}
%
\item This condition holds by the following sets of scheme
  generalisations: for the upper bound, $\forall u \in M_c,
  \exists v \in M_s . u \preceq_2 v$: 
  \begin{align}
    \begin{array}{l}
     \vtwoh{0}{-1} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{0}{1} \preceq_2 \vtwoh{0}{\infty} \\[0.5em]
     \vtwoh{-1}{0} \preceq_2 \vtwoh{\infty}{0} \quad
     \vtwoh{1}{0} \preceq_2 \vtwoh{\infty}{0}
     \end{array}
  \label{eq:check-example-upper}
  \end{align}
  % 
  and for the lower bound: $\forall v \in M_s, \exists u \in M_c . u
  \preceq_n v$: 
  \begin{align}
    \hspace{-1em}
    \begin{array}{l}
     \vtwoh{-1}{0} \preceq_2 \vtwoh{-1}{\infty} \quad
     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{1}{0} \preceq_2 \vtwoh{1}{\infty}  \\
     \vtwoh{0}{-1}  \preceq_2 \vtwoh{\infty}{-1} \quad
     \vtwoh{0}{0}  \preceq_2 \vtwoh{\infty}{0} \quad
     \vtwoh{0}{1}  \preceq_2 \vtwoh{\infty}{1}
     \end{array}
    \label{eq:check-example-lower}
  \end{align}
  Since both the consistency predicate is satisfied, the array
  accesses flowing to line three in the stencil computation are valid
  with respect to the {\textcolor{darkgreen}{\texttt{fivepoint}}} specification.
\end{enumerate}
\vspace{-0.7em}
\end{example}
\begin{example} 
We consider now an erroneous user modification to $l3$ with an
additional term \texttt{b(i+2,j+2)} such that $l3'$ is:
%
\begin{minted}[firstnumber=3]{fortran}
a(i,j) = b(i,j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
\end{minted}
%
The checking procedure proceeds in the same way as the first three
steps of the previous example, except that the result of
$\mathit{modelise}(l3',\texttt{b})$ generates an additional scheme
$\vtwoh{2}{2}$. Since the spec model has not changed and the code
model is has only the additional scheme $\vtwoh{2}{2}$, the generalisations for
the lower bound \eqref{eq:check-example-lower} 
of step 4 above still holds since its consistency condition
universally quantifies over the \emph{spec model}. However, for
the upper bound, the consistency condition universally quantifies
over scheme $u$ of the \emph{code model} for which there exists
a scheme $v \in M_s$ such that $u \preceq_2 v$. Thus, we require a $v
\in M_s$ such that $\vtwoh{2}{2} \preceq_2 v$ but no such generalising
scheme existing in the spec model. Hence the model fails to hold as an upper bound.

@@

As noted, the {\textcolor{darkgreen}{\term{fivepoint}}} region
 above is more accepting than that used in
Section~\ref{sec:lang}. This introduces a new notion of approximation
beyond what is given by the syntactic theory of sub-specifications
$<:$ in Section~\ref{sec:eqs}. Consistency leads to a notion of
approximation that is a \textit{semantic sub-specification} relation:
%using upper and lower bounds because the in both cases the specifications are
%used to generate \emph{exact} models.

@@

We first examined how frequently stencil computations occur in our
corpus. This was carried out by running our implementation in 
whole-code-base inference mode. We define a \emph{potential} statement as one whose left-hand
side is an array subscript on neighbourhood indices. These statements
are the inputs to Step 2 of the Inference process described in 
Section~\ref{subsec:inference}. Each potential statement for which we
eventually generate some specification is also counted as
an \emph{actual} statement.

>>>>

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{color}
\usepackage{graphics}
\usepackage{fancyvrb}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{minted}
\usepackage{breqn}
\usepackage{enumitem}
%\usepackage{subcaption}
\usepackage{siunitx} % For pretty-printing numeric values and SI units
                     % of measure. e.g., the tabular column type S is
                     % used to print nice-looking tables of numbers.
\sisetup{ % defaults
  group-separator={,},
  group-minimum-digits={3},
  output-decimal-marker={.},
  table-format = 6
}

\usepackage{tikz}
\usepackage{xypic}
\usepackage{natbib}

@@

\newtheorem{lemma}[block]{Lemma}
\newtheorem{proposition}[block]{Proposition}

@@

\title{Verification of Stencil Computations with Spatial
  Specifications}
\titlebanner{DRAFT -- Do not distribute}
\authorinfo{}{}{}
\authorinfo{Dominic Orchard \and Mistral Contrastin
\and Matthew Danish \and Andrew Rice}{University of Cambridge}{}

@@

In this paper, we make the following contributions:
%
\begin{itemize}[leftmargin=0.8em]
\item we introduce a specification language for stencils 
  that captures many common forms of data access pattern
  (\S~\ref{sec:lang});

\item we detail inference, checking, and specification synthesis 
algorithms for our language (\S~\ref{sec:analysis}), derived from
a denotational model for the language (\S~\ref{sec:semantics});

@@

Our specification language is based on the hypothesis
that most forms of array access in numerical code have
a fixed, statically-determined access pattern. For example, the
\emph{five-point stencil} on a two-dimensional array reads from array
indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
for all $i, j$ within the inner boundary of the array (to avoid
out-of-bounds access at the edges). We revisit this hypothesis
in \S~\ref{sec:evaluation} with the inference of
such regular stencil patterns on our corpus of numerical programs.

\S~\ref{sec:syntax} outlines the syntax.
 \S~\ref{sec:eqs} provides an equational
theory for specifications and a notion of
sub-specification.  \S~\ref{sec:semantics}
gives a semantic model via a set-based interpretation, for which the
equality and sub-specification relation are proven sound.

@@

The last six equations (from (\textsc{\term{+}idem}) inclusive), explain
the algebraic behaviour of \term{+} and \term{*}.  Together, we see
that \term{+} is an associative, commutative, idempotent binary
operator that distributes with \term{*}, which is associative and
commutative. This distribution of \term{*} over \term{+} is key as it
is used internally to give a normal form for stencil specifications
akin to Disjunctive Normal Form (DNF) (where \term{+} is taken as
``disjunction'' and \term{*} as the ``conjunction''). We revisit this
normalisation in \S~\ref{sec:inf-step4}.

The equational theory provides not only an axiomatic semantics
for the language, which the user can use to rewrite their
specification, but it is also used extensively in the inference
procedure (\S~\ref{subsec:inference}) to simplify specifications.

@@

We define a denotational model of the semantics of our 
stencil specification language. This model has a number of purposes: (1) it is used in the inference,
checking, and synthesis algorithms (\S~\ref{sec:analysis}); (2)
justifies the equational
theory and approximations of the previous section; and (3) can be used
to guide correct implementations. The model domain is over sets of vectors which we call \emph{index
  schemes}, which we introduce first before the model in \S~\ref{subsec:model}.

@@

\begin{definition} \textsf{Mult}
and \textsf{Approx} are parametric labelled variant types
with injection function given by their definition:
%
\begin{align*}
\textsf{Mult} \, a & = \textsf{mult} \, a \,\mid\, \textsf{only} \, a \\
\textsf{Approx} \, a & = \textsf{exact} \, a \,\mid\, \textsf{lower} \, a \,\mid\,
\textsf{upper} \, a \,\mid\, \textsf{both} \, a \, a
\end{align*}
\ie{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$, and so
on. The $\textsf{Mult}$ type corresponds to the presence or absence of the
\term{readOnce} modifier as shown in Figure~\ref{fig:spatial-model}. 
The $\textsf{Approx}$ type corresponds to the presence or absence of
the spatial approximation modifier, with \textsf{exact} when there is
no such modifier and \textsf{lower} and \textsf{upper} for
\term{atLeast} and \term{atMost}. Later, it will be useful to model
pairs of lower and upper bounds which is provided by \textsf{both} 
(but which isn't used here). These are used for labelled models, and other components later in
\S~\ref{sec:analysis}.% with additional information.
\label{def:mult-and-approx}
\end{definition}
\vspace{-0.5em}

@@

\noindent
We briefly sketch the algorithms for inferring and synthesising
specifications (\S~\ref{subsec:inference}), for checking code
against specifications (\S~\ref{subsec:checking}). These utilise the semantic model of
\S~\ref{sec:semantics}.

@@

\begin{definition}\label{def:span-coalesc} Span
  coalescing $\bullet$ is defined for spans contiguous in the first dimension:
%%
\begin{align*}
\vect{x} \bullet \vect{y}
= &
\begin{cases}
(\vect{x}^L, \vect{y}^U) & \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
\vect{x}_{2:n} = \vect{y}_{2:n}) \\
\bot  & \textit{otherwise}
\end{cases}
% $m$-dimensions into coalesced spans which vary in
% $m+1$-dimensions.
\end{align*}
%%
\end{definition}
\noindent
To coalesce regions which are contiguous in other dimensions
other than the first, we generate all rotations of the spans in
$n$-dimensional space by permutation on spans, and fold together contiguous
regions using $\bullet$ (above). This 
generates all possible ways of coalescing regions. For our example,
the unit spans in Figure~\ref{fig:inference-steps-informal}(a)
are coalesced into the contiguous spans in
Figure~\ref{fig:inference-steps-informal}(b).
This procedure is iterated until a fixed point is reached. In this
example this is reached in the first step. 
 Then any spans that are completely contained by any other span are deleted,
leaving a minimal set of spans (which may overlap, but none of which
fully contains another), Figure~\ref{fig:inference-steps-informal}(c).
The final spans are used to synthesise concrete stencil specification
syntax (\S~\ref{sec:inf-step4}), where
each span becomes a product \texttt{*} of regions which are 
combined by the sum \texttt{+}.

@@

In checking mode, the access pattern of a stencil computation in the
source language is verified against a specification. The model of
\S~\ref{sec:semantics} again aids this process.  Checking
proceeds by generating a model $M$ from a specification and generating a set $N$ of indexing
schemes from the source code. The source code is then valid with
respect to the specification if $N$ is \emph{consistent} with $M$. The
notion of consistency is built out of index scheme 
\emph{generalisation} (Sect.~\ref{subsec:schemes-and-gen}). 
%Consistency is defined in Figure~\ref{fig:consistency},
We first set up some intermediate definitions.% before unpacking its definition.

@@

\begin{example}
We give an extended example of verifying a stencil computation against
its specification. We consider a more general, but still consistent, version of the \textit{five-point}
stencil specification inferred previously in
\S~\ref{subsec:inference}:
%
\begin{minted}[breakindent=8.1em]{fortran}
!= region :: fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
!= stencil readOnce, fivepoint :: b
a(i,j) = b(i,j-1)+b(i,j)+b(i,j+1)+b(i-1,j)+b(i+1,j)
\end{minted}
%
We refer to the assignment on line three as $l3$ for brevity.

Verification of the code against the spec is then given by 
$check(({\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}),l3,\texttt{b})$. 
We explain the steps.
\begin{enumerate}  
%
\item Generate the model from the specification as in
  \S~\ref{sec:semantics}. 
%
\begin{align*}
 M_s= \{  \vtwo{-1}{\infty},
          \vtwo{0}{\infty},
          \vtwo{1}{\infty},
          \vtwo{\infty}{-1},
          \vtwo{\infty}{0},
        \vtwo{\infty}{1} \}  \\[0.4em]
\interp{{\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}}_2 =
  \textsf{once}(\textsf{exact} (M_s))
\end{align*}
%
\item Obtained the code model via $\textit{modelise}$, giving:
%
\begin{align*}
M_c = \{
    \vtwo{0}{-1},
    \vtwo{0}{0},
    \vtwo{0}{1},
    \vtwo{-1}{0},
    \vtwo{1}{0}
  \}
\\[0.4em]
\textit{modelise}(l3, \texttt{b}) = \textsf{once} (M_c)
\end{align*}
%
\item Consistency $cons_2(\textsf{once}(\textsf{exact}(M_s)),
  \textsf{once}(M_c))$ holds when 
  $\textit{cons}_2(\textsf{mult}(\textsf{exact}(M_s)),\textsf{mult}(M_c))$ (Fig.~\ref{fig:consistency}) 
  which holds when: 
  \begin{align*}
     & \textit{cons}_2(\textsf{mult}(\textsf{up}(M_s)),\textsf{mult}(M_c)) \\
   \wedge \; & \textit{cons}_2(\textsf{mult}(\textsf{low}(M_s),\textsf{mult}(M_c))
  \end{align*}
%
\item This condition holds by the following sets of scheme
  generalisations: for the upper bound, $\forall u \in M_c,
  \exists v \in M_s . u \preceq_2 v$: 
  \begin{align}
    \begin{array}{l}
     \vtwoh{0}{-1} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{0}{1} \preceq_2 \vtwoh{0}{\infty} \\[0.5em]
     \vtwoh{-1}{0} \preceq_2 \vtwoh{\infty}{0} \quad
     \vtwoh{1}{0} \preceq_2 \vtwoh{\infty}{0}
     \end{array}
  \label{eq:check-example-upper}
  \end{align}
  % 
  and for the lower bound: $\forall v \in M_s, \exists u \in M_c . u
  \preceq_n v$: 
  \begin{align}
    \hspace{-1em}
    \begin{array}{l}
     \vtwoh{-1}{0} \preceq_2 \vtwoh{-1}{\infty} \quad
     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{1}{0} \preceq_2 \vtwoh{1}{\infty}  \\
     \vtwoh{0}{-1}  \preceq_2 \vtwoh{\infty}{-1} \quad
     \vtwoh{0}{0}  \preceq_2 \vtwoh{\infty}{0} \quad
     \vtwoh{0}{1}  \preceq_2 \vtwoh{\infty}{1}
     \end{array}
    \label{eq:check-example-lower}
  \end{align}
  Since both the consistency predicate is satisfied, the array
  accesses flowing to line three in the stencil computation are valid
  with respect to the {\textcolor{darkgreen}{\texttt{fivepoint}}} specification.
\end{enumerate}
\vspace{-0.7em}
\end{example}
\begin{example} 
We consider now an erroneous user modification to $l3$ with an
additional term \texttt{b(i+2,j+2)} such that $l3'$ is:
%
\begin{minted}[firstnumber=3]{fortran}
a(i,j) = b(i,j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
\end{minted}
%
The checking procedure proceeds in the same way as the first three
steps of the previous example, except that the result of
$\mathit{modelise}(l3',\texttt{b})$ generates an additional scheme
$\vtwoh{2}{2}$. Since the spec model has not changed and the code
model is has only the additional scheme $\vtwoh{2}{2}$, the generalisations for
the lower bound \eqref{eq:check-example-lower} 
of step 4 above still holds since its consistency condition
universally quantifies over the \emph{spec model}. However, for
the upper bound, the consistency condition universally quantifies
over scheme $u$ of the \emph{code model} for which there exists
a scheme $v \in M_s$ such that $u \preceq_2 v$. Thus, we require a $v
\in M_s$ such that $\vtwoh{2}{2} \preceq_2 v$ but no such generalising
scheme existing in the spec model. Hence the model fails to hold as an upper bound.

@@

As noted, the {\textcolor{darkgreen}{\term{fivepoint}}} region
 above is more accepting than that used in
\S~\ref{sec:lang}. This introduces a new notion of approximation
beyond what is given by the syntactic theory of sub-specifications
$<:$ in \S~\ref{sec:eqs}. Consistency leads to a notion of
approximation that is a \textit{semantic sub-specification} relation:
%using upper and lower bounds because the in both cases the specifications are
%used to generate \emph{exact} models.

@@

We first examined how frequently stencil computations occur in our
corpus. This was carried out by running our implementation in 
whole-code-base inference mode. We define a \emph{potential} statement as one whose left-hand
side is an array subscript on neighbourhood indices. These statements
are the inputs to Step 2 of the Inference process described in 
\S~\ref{subsec:inference}. Each potential statement for which we
eventually generate some specification is also counted as
an \emph{actual} statement.
